Embedding to Reference t-SNE Space Addresses Batch Effects in Single-Cell Classification

### 2021-05-13 paper-revision/response-to-reviewers.tex

\documentclass{article}

\usepackage[T1]{fontenc}
\usepackage{natbib}
\usepackage{todonotes}
\usepackage{gensymb}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage[top=1in,bottom=1in,left=1in,right=1in]{geometry}
\usepackage{alltt}
\usepackage{framed}
\usepackage{xcolor}
\usepackage{comment}
\usepackage{graphicx}

\usepackage{xr}[]
\externaldocument{policar2019-tsne-embedding}
\usepackage{url}

\renewcommand{\thetable}{\Alph{table}}
\renewcommand{\thefigure}{\Alph{figure}}

\newcommand{\beq}{\begin{equation}}
\newcommand{\eeq}{\end{equation}}
\newcommand{\beqs}{\begin{equation}}
\newcommand{\eeqs}{\end{equation}}
\newcommand{\mx}[1]{\ensuremath{\boldsymbol{\mathrm{{#1}}}}}   % Simple matrix format.

% \newcommand{\review}[1]{\noindent{{\textbf{\textit{#1}}}}\vspace{3mm}}
% \definecolor{colframe}{rgb}{0.01,0.199,0.1}

\newcounter{rtaskno}
\newcommand{\rtask}[1]{\refstepcounter{rtaskno}\label{#1}}

\newcommand{\reviewc}[2]{\begin{leftbar}\noindent\rtask{#1}\ref{#1}. #2\end{leftbar}}
\newcommand{\review}[1]{\begin{leftbar}\noindent #1\end{leftbar}}
\newcommand{\answer}[1]{\noindent #1\\[0mm]}
\newcommand{\copyquote}[1]{\vspace{2mm} {\color{blue} #1}\vspace{2mm}}

\newlength{\leftbarwidth}
\setlength{\leftbarwidth}{3pt}
\newlength{\leftbarsep}
\setlength{\leftbarsep}{10pt}
\renewenvironment{leftbar}{%
    \def\FrameCommand{{\color{black}{\vrule width \leftbarwidth\relax\hspace {\leftbarsep}}}}%
    \MakeFramed {\advance \hsize -\width \FrameRestore }%
}{\endMakeFramed}

% \usepackage[colorlinks=True,citecolor=black]{hyperref}

\begin{document}

\noindent{\bf Poli\v{c}ar, Stra\v{z}ar \& Zupan: Embedding to Reference t-SNE Space Addresses Batch Effects in Single-Cell Classification}

\vspace{5mm}
\noindent {\bf RESPONSE TO THE REVIEWERS}

\vspace{5mm}

\noindent Dear Editor and the Reviewers, \\

\vspace{3mm}

\noindent we thank for the comprehensive feedback that helped to improve the
manuscript, sufficient time to implement the changes, and the permission to
submit a revised version.

\vspace{3mm}

All of the reviewer remarks are addressed below with reference to the
corresponding changes in the manuscript.  In case of overlapping remarks from
different reviewers, we refer the reader to the relevant enumerated answer. A
differential document is provided to facilitate the comparison with the initial
submission. The changes to the data, code and the instructions to reproduce all
the experiments are available in the repository:
\url{https://github.com/biolab/tsne-embedding}.

\vspace{3mm}

\begin{flushright}
\noindent Kind regards, \\ 
\vspace{2mm}
Pavlin Poli\v{c}ar \\
Martin Stra\v{z}ar \\
Bla\v{z} Zupan \\ 
\vspace{2mm}
University of Ljubljana, Slovenia 
\end{flushright}

\vspace{10mm}

\subsection*{Reviewer \#1}

\reviewc{r1q1}{Most important is to add in theoretical part investigation about which dataset can be a reference dataset. It is necessary because it is crucial in estimation of level of trust worth.
Some information about it is provided in testing part but only in form of general statement on specific case (based on some data sets), but it is necessary to have a procedure/test when given data set is a good reference data set.}

\answer{
	We thank the reviewer for this comment and have expanded our discussion of potential failure cases in the ``Experiments and Discussion'' section. We have added an example of a failure case and have pointed out that care must be taken when using our procedure.

	A theoretical investigation of which data set can be used as a reference is beyond the scope of this paper. Conceptually, our approach for classification is very similar to a $k$-nearest neighbor classifier. It would be very difficult to theoretically quantify which reference data sets would be appropriate for such classifiers.
}

\reviewc{r1q2}{Page 1
Bibliography:
positions 3 and 4 is not about PCA and MDS. Please provide appropriate citations for PCA and MDS.}

\answer{We thank the reviewer for catching this mistake. We have fixed these citations.}

\reviewc{r1q3}{
Page 6
How do you select L and $\sigma_{i,l}$?
}

\answer{
	This isn't part of our work, and has been moved to the supplement. Refer to the original publication for a thorough comparison. We moved this section to the appendix.
}

\reviewc{r1q4}{page 9
The "I(.)" in Eq. 11 is not defined.}

\answer{
	We've added a short sentence, explaining that this is the indicator function. We moved this section to the appendix.
}

\reviewc{r1q5}{Please precise what do you mean by the right side of Eq. 12 (the meaning of brackets and "|").}

\answer{We have modified the text so that the equations in this section are clearer.}

\reviewc{r1q6}{Generally subsection 3.3 is not as clear as it should be. Please rewrite it.}

\answer{We agree that this subsection could be more clear. We have modified some parts of it and added sentences which should clarify the gene selection procedure.}

\reviewc{r1q7}{Figs 1-4
It would be much more readable when Figures would be bigger. Please provide subfigures vertically, not horizontally.}

\answer{}

\reviewc{r1q8}{GENERAL statement: provide better quality of all Figures (better resolution, now we see large pixels too early...)}

\answer{}

\reviewc{r1q9}{Fig 5
Please modify comparison presented in Fig 5.
Please provide 4 subfigures:
- t-SNE without PCA and without multiscale
- t-SNE with PCA and without multiscale
- t-SNE without PCA and with multiscale
- t-SNE with PCA and with multiscale
only in such case we can be sure what is real source of success.
It would be fine to see such comparison on (at least) two datasets. (!)}

\answer{
	This isn't part of our work, and has been moved to the supplement. Refer to the original publication for a thorough compoarison.
}

\reviewc{r1q10}{Fig 6
Authors claim that in case of sub-figure c all genes may lead to under-clustering but the figures b and c are very similar in the number of clusters.}

\answer{
	How to select the number of genes is pretty irrelevant to our work. This is a question of single-cell processing pipeline, which we do not care about in this work.
}

\subsection*{Reviewer \#2}

\reviewc{r2q1}{Since the focus, from an application point of view, is on bioinformatics and specifically on batch effects, I think that a more detailed description of this phenomenon should be reported in the introduction. }

\answer{
	We've addressed this in related work, where we've added a paragraph dedicated to batch effects.
}

\reviewc{r2q2}{The structure of the paper can significantly be improved. In particular:

- A section devoted to the description of related work should be introduced, reporting existing work on both the application and methodological viewpoints. In this respect, some descriptions currently reported in the experiments sections (see, for example, the description of scMap-cluster) should be moved in this section. Moreover, I actually can see a strong similarity with transfer learning methods, specifically with those focusing on domain adaptation. I suggest the authors to consider some relevant recent works in this field in their discussion (see, for example, the survey 10.1186/s40537-016-0043-6), also specifically focusing on bioinformatics (see, for example, the work 10.1093/bioinformatics/btz781).

- Related to the previous comment, some details related to existing approaches should be removed (or strongly simplified) from the section devoted to experiments, and moved to a new section devoted to related work.

- Some parameter values reported in the methods section should be moved in the experiments, since they are part of the experimental setting.}

\answer{We thank the reviewer for this comment. We have restructured the paper.

	We have added a ``Related work'' section, where we describe batch-effects in more detail. We have also moved the description of scMap-Cluster from the ``Experiments'' section to this section. While it is true that these approaches are, in essence, transfer learning, the data sets produced by single-cell RNA-seq are sufficiently different from data sets from other fields of machine learning that they requires specialized methods, which we describe in the ``Related work'' section.
}

\reviewc{r2q3}{The current paper is an extension of a conference paper submitted to the conference Discovery Science 2019. The authors should explicitly provide in the text the actual novel contributions with respect to the conference version.}

\answer{}

\reviewc{r2q4}{The authors should better emphasize the contribution from a methodological viewpoint. Indeed, without a proper emphasis on this aspect, the paper seems to be the application of an existing tool (namely, openTSNE) to a new dataset. If the contribution from a methodological viewpoint is actually limited, the authors should still put some effort in emphasizing the advantages of this approach in this specific domain.}

\answer{
	We have changed the working in ``Availability and Implementation'' section, which emphasizes that the extension we describe in this paper was developed by us and later incorporated into \textsf{openTSNE}. As this is the only mention of \textsf{openTSNE} in the manuscript, we believe there can be no further confusion in this respect.
}

\reviewc{r2q5}{PCA-based initialization seems to help a lot. I would appreciate to see a comparison of the results with those achievable with the simple application of PCA with just 2 principal components.}

\answer{
	This isn't part of our work, and has been moved to the supplement. Please, refer to the original publication for a thorough comparison. However, we would also refer the reviewer to a recent comment by Kobak and Linderman ``Initialization is critical for preserving global data structure in both t-SNE and UMAP'', \textit{Nature Biotechnology}, 2021, which demonstrates that good initialization schemes are important for both t-SNE and UMAP.
}

\reviewc{r2q6}{At the beginning of Section 3.2, a quite long preprocessing pipeline is shown. Actually, I would like to understand the underline motivations behind that. Few additional sentences would help in this respect.}

\answer{
	This is a question of single-cell processing pipeline, which is not the focus of our work. We provide a reference to the original publication presenting the pipeline in the text.
}

\subsection*{Reviewer \#3}

\reviewc{r3q1}{The research heavily relies on the ref. [15]. Although the authors claimed that they provide both practical and theoretical grounded implementation, the contribution in terms of theoretical justification is not convincing enough. The equations, given in Section 2, closely correlate with those provided in [15], except for two small modifications - replacing x and y with v and w. }

\answer{
	We agree that the original manuscript included sections that are important to the single-cell analysis pipeline which are perhaps not the most relevant to our particular contribution i.e. the extension of t-SNE to allow embedding new data points into existing embeddings. For this particular task, the details of the single-cell analysis are not crucial. While it is true that our single-cell pipeline relies on the work from [15], we have moved most of these sections to the appendix, as they are not directly related to our work. We believe that this improves the readability of our paper. However, the work done in [15] introduced ideas which are now considered standard practice for t-SNE visualizations, so a brief review of the most important concepts is kept in Section 3.1.

	Regarding the contribution in terms of theoretical justification: the extension is indeed very minor, and requires changing only two variables, which results in different interactions being calculated. However, additional minor but important changes need to made, including the omission of the symmetrization step and a minor change in the gradient calculation.

	Perhaps more important is the practical aspect of our work i.e. how do we get this to work. While the equations change only slightly, a straightforward application of our approach using the currently accepted t-SNE parameter settings will result in wildly unstable optimization, so special care must be taken to mitigate this erratic behaviour. We describe this in the last paragraph of Section 3.2, where we emphasize the importance of lowering the learning rate and introducing optimization tricks e.g. gradient clipping. This is especially important for the approximation schemes used today by virtually all t-SNE implementations, as a poor parameter setting will result in points ``shooting-off'' far away from the reference points and will dramatically impact the runtime of these approximation schemes. Without these modifications, the t-SNE transform functionality becomes unusable in virtually all settings.
}

\reviewc{r3q2}{From the Abstract, it is not immediately clear what the reference and secondary datasets are. The authors explained it later in the main body, but it seems reasonable to introduce the terms earlier to keep readers interested. Further, the authors claim that they show "the utility of the proposed approach by analyzing 6 recently published ... datasets … with up to tens of thousands of cells and thousands of genes ..."  and so on. Mentioning more specific quantitative results with regard to particular datasets, in comparison with the known results, would be desirable. }

\answer{
	We have slightly reworded the abstract to indicate that the secondary data set contains new, unseen points.

	We disagree that mentioning more qualitative results would improve this manuscript. There are no robust, agreed-up baselines in the single-cell literature for batch effect removal in general. Only a handful of methods for cross-data set cell-type classification methods have been proposed, and none of them are directly related to our approach. This manuscript highlights that a simple extension to a visualization technique, t-SNE, performs competitively to specialized machine learning and bioinformatics methods for cell-type classification. These methods are not directly comparable, and our aim is not to improve the accuracy of such approaches. We have, however, reworded the last paragraph of Section 4.6 to emphasize this point.
}

\reviewc{r3q3}{The Results and Discussion section is difficult to follow. Please, if possible, consider splitting it in subsections.}

\answer{
	We thank the reviewer for this comment and we agree that splitting the section into subsections would improve the readability of the manuscript. We have split up the ``Results and Discussion'' section into several subsections.
}

\reviewc{r3q4}{Consider replacing the legends in Fig. 1 with the names or descriptions of the datasets to increase the readability. }

\answer{
	The legends of Figure 1 show the names of the data sets, or rather, their original publication. The figure demonstrates that two data sets with shared cell types cluster primarily by data set and not by cell-type. In this instance one would reasonably assume and even hope that the samples would cluster by cell-type and that the data set source would not interfere with the visualization. We feel that a description of the data sets is unnecessary here, as it does not help illustrate our point, and would needlessly burden the reader with biological details.
}

\reviewc{r3q5}{Please extend Section 3.5, which includes only two sentences, with more details of implementation or consider moving this text into Section 3.4 (possibly at the beginning or at the end).}

\answer{We agree that this is not the most suitable location for this information, and we have moved it to the end of the manuscript, after the conclusion and before the acknowledgments, as is standard practice in many journals.}

\reviewc{r3q6}{Please update the preprint [15] with the published manuscript.}

\answer{The thank the reviewer for the reminder. We have updated the citation.}

\end{document}

### 2021-05-13 paper-revision/diffbase.tex
Add revision files

\begin{document}
%
\title{Embedding to Reference t-SNE Space Addresses Batch Effects in Single-Cell Classification}
%
\titlerunning{t-SNE Embedding and Batch Effects}
%
\author{Pavlin G. Poli\v{c}ar\inst{1} \and
Martin Stra\v{z}ar\inst{1} \and
Bla\v{z} Zupan\inst{1,2}}
%
\authorrunning{P. G. Poli\v{c}ar, M. Stra\v{z}ar, and B. Zupan}
%
\institute{University of Ljubljana, SI-1000 Ljubljana, Slovenia
\email{\{pavlin.policar,martin.strazar,blaz.zupan\}@fri.uni-lj.si}\\[2pt]
\and
Baylor College of Medicine, Houston, TX 77030, USA
}

@@

Dimensionality reduction techniques, such as t-SNE, can construct informative
visualizations of high-dimensional data. When working with multiple data
sets, a straightforward application of these methods often fails; instead of
revealing underlying classes, the resulting visualizations expose data
set-specific clusters. To circumvent these batch effects, we propose an
embedding procedure that uses a t-SNE visualization constructed on a
reference data set as a scaffold for embedding new data points. The new,
secondary data is embedded one data point at the time. This prevents any
interactions between instances in the secondary data and implicitly mitigates
batch effects. We demonstrate the utility of this approach by analyzing six
recently published single-cell gene expression data sets with up to tens of
thousands of cells and thousands of genes. The batch effects are particularly
strong as the used data comes from different institutions and was obtained
using different experimental protocols. The visualizations constructed by our
proposed approach are cleared of batch effects, and the cells from secondary
data sets correctly co-cluster with cells of the same type from the primary
data.

\keywords{Batch effects \and Embedding \and t-SNE \and Visualization \and Single-Cell Transcriptomics \and Data Integration \and Domain Adaptation.}
\end{abstract}

@@

Massive, multi-variate biological data sets often suffer from these
source-specific biases.  The focus of this work is single-cell genomics, a
domain that was selected due high biomedical relevance and and abundance of
recently published data. Single-cell RNA sequencing (scRNA-seq) data sets are
the result of isolating RNA molecules from individual cells, which serve as an
estimate of the expression of cell's genes.  The studies can exceed thousands
of cells and tens of thousands of genes, and typically start with cell type
analysis. Here, it is expected that cells of the same type would cluster
together in two-dimensional data visualization~\cite{seurat}. For instance,
Fig.~\ref{fig:batch_effect}.a shows t-SNE embedded data from mouse brain cells
originating from the visual cortex~\cite{hrvatin2018} and the
hypothalamus~\cite{chen2017}. The figure reveals distinct clusters but also
separates the data from the two brain regions. These two regions share the same
cell types and --- contrary to the depiction in Fig.~\ref{fig:batch_effect}.a
--- we would expect the data points from the two studies to overlap. Batch
effects similarly prohibit the utility of t-SNE in the exploration of
pancreatic cells in Fig.~\ref{fig:batch_effect}.b, which renders the data from
a pancreatic cell atlas~\cite{baron2016} and similarly-typed cells from
diabetic patients~\cite{xin2016}. Just like with data from brain cells,
pancreatic cells cluster primarily by data source, again resulting in a
visualization driven by batch effects.

\begin{figure}[htbp]
  \includegraphics[width=\textwidth]{figures/batch_effect.pdf}
  \caption{Batch effects are a driving factor of variation between the data
  sets. We depict a t-SNE visualization of two pairs of data sets. In each
  pair, the data sets share cell types, so we would expect cells
  from the reference data (blue) to mix with the cells in a secondary data
  sets (orange). Instead, t-SNE clusters data according to the
  data source.}
  \label{fig:batch_effect}
\end{figure}

Current solutions to embedding the data from various data sources address the
batch effect problems up-front. The data is typically preprocessed and
transformed such that the batch effects are explicitly removed. Recently
proposed procedures for batch effect removal include canonical correlation
analysis~\cite{cca} and mutual nearest-neighbors~\cite{mnn,seurat}.  In these
works, batch effects are deemed removed when cells from different sources
exhibit good mixing in a t-SNE visualization.  The elimination of batch effects
may require aggressive data preprocessing which may blur the boundaries between
cell types. Another problem is also the inclusion of any new data, for which
the entire analysis pipeline must be rerun, usually resulting in a different
layout and clusters that have little resemblance to original visualization and
thus require reinterpretation.

We propose a direct solution of rendering t-SNE visualizations to address
batch effects. Our approach treats one of the data sets as a {\em reference}
and embeds the cells from another, {\em secondary data set} to a joint
low-dimensional space.  We construct a t\nobreakdash -SNE embedding using the
reference data set, which is then used as a scaffold to embed the
secondary data. The key idea underpinning our approach is that secondary data
points are embedded independently of one another.

@@

\begin{figure}[htb]
  \includegraphics[width=\textwidth]{figures/transform_brain.pdf}
  \caption{A two-dimensional embedding of a reference containing brain cells
  (a) and the corresponding mapping of secondary data containing hypothalamic
  cells (b).  The majority of hypothalamic cells were mapped to their
  corresponding reference cluster. For instance, astrocyte cells marked with
  red on the right were mapped to an oval cluster of same-typed cells denoted
  with the same color in the visualization on the left.}
  \label{fig:transform_brain}
\end{figure}

The proposed solution implements a mapping of new data into an existing
t-SNE visualization. While the utility of such an algorithm was already hinted
at in recent publication~\cite{art_of_using_tsne}, here we provide its
practical and theoretically-grounded implementation. Considering the abundance
of recent publications on batch effect removal, we present surprising evidence
that a computationally more direct and principled embedding procedure solves
the batch effects problem when constructing interpretable visualizations from
different data sources.

@@

We describe an end-to-end pipeline that uses fixed t-SNE coordinates as a
scaffold for embedding new (secondary) data, enabling joint visualization of
multiple data sources while mitigating batch effects. Our proposed approach
starts by using t\nobreakdash -SNE to embed a reference data set, with the aim
of constructing a two-dimensional visualization to facilitate interpretation
and cluster classification. Then, the placement of each new sample is optimized
independently via the t\nobreakdash -SNE loss function. Independent treatment
of each data instance from a secondary data set disregards any interactions
present in that data set, and prevents the formation of clusters that would be
specific to the secondary data. Below, we start with a summary of t-SNE and its
extensions (Sec.~\ref{sec:tsne}, introducing the relevant notation, upon which
we base our secondary data embedding approach (Sec.~\ref{sec:transfer}).

\subsection{Data embedding by t-SNE and its extensions\label{sec:tsne}}

@@

\subsection{Adding new data points to reference embedding\label{sec:transfer}}

@@

In the optimization step, we refine point positions using batch gradient
descent. We use an adaptive learning rate scheme with momentum to speed up the
convergence, as proposed by Jacobs~\cite{momentum,bh_tsne}. We run gradient
descent with momentum $\alpha$ of $0.8$ for $250$ iterations, where the
optimization converged in all our experiments. The time complexity needed to
evaluate the gradients in Eq.~\ref{eq:gradient} is $\mathcal{O}(N \cdot M)$,
however, by adapting the same polynomial interpolation based approximation,
this is reduced to $\mathcal{O}(\max \{ N, M \})$.

Special care must be taken to reduce the learning rate $\eta$ as the default
value in most implementations ($\eta = 200$) may cause points to ``shoot off''
from the reference embedding. This phenomenon is caused due to the embedding to
a previously defined t-SNE space, where the distances between data points and
corresponding gradients of the optimization function may be quite large. When
running standard t-SNE, points are initialized and scaled to have variance
0.0001. The resulting gradients tend to be very small during the initial phase,
resulting in stable convergence. When embedding new samples, the span of the
embedding is much larger, resulting in substantially larger gradients, and the
default learning rate causes points to move very far from the reference
embedding. In our experiments, we found that decreasing the learning rate to
$\eta \sim 0.1$ produces stable solutions. This is especially important when
using the interpolation-based approximation, which places a grid of
interpolation points over the embedding space, where the number of grid points
is determined by the span of the embedding. Clearly, if even one point ``shoots
off'' far from the embedding, the number of required grid points may grow
dramatically, increasing the runtime substantially. The reduced learning rate
suppresses this issue, and does not slow the convergence because of the
adaptive learning rate scheme, provided the optimization is run for a
sufficient number of steps.

@@

\vspace{-3mm}
\begin{table}[ht]
\begin{center}
\setlength\tabcolsep{4pt}
\begin{tabular}{l c c r c c}
\toprule
Study & Organism/Tissue & Protocol & Cells & Cell Types & Sparsity (\%) \\
\midrule
Hrvatin \etal & \multirow{2}{*}{mouse brain} & inDrop & 48,266 & 9 & 94 \\
Chen \etal & & Drop-seq & 14,437 & 6 & 93 \\[5pt]
Baron \etal & \multirow{2}{*}{human pancreas} & inDrop & 8,569 & 9 & 91 \\
Xin \etal & & SMARTer & 1,492 & 4 & 86 \\[5pt]
Macosko \etal & \multirow{2}{*}{mouse retina} & Drop-seq & 44,808 & 12 & 97 \\
Shekhar \etal & & Drop-seq & 27,499 & 5 & 96 \\
\bottomrule
\end{tabular}
\end{center}
\caption{Data sets used in our experiments. The first data set in a pair
(Hrvatin \etal, Baron \etal, and Macoscko \etal) was used as a reference.  We
relied on the quality control and annotations from the original publication
and report the humber of cell types after preprocessinng.  The cell
annotations were made consistent to annotations from the Cell
Ontology~\cite{cell_ontology}. Notice that different RNA sequencing protocols
were used to estimate gene expressions.}
\label{tab:data sets}
\end{table}

\subsection{Single-cell data preprocessing pipeline}

>>>>

\begin{document}
%
\title{Embedding to Reference t-SNE Space Addresses Batch Effects in Single-Cell Classification}
%
\titlerunning{t-SNE Embedding and Batch Effects}
%
\author{Pavlin G. Poli\v{c}ar\inst{1} \and
Martin Stra\v{z}ar\inst{1} \and
Bla\v{z} Zupan\inst{1,2}}
%
\authorrunning{P. G. Poli\v{c}ar, M. Stra\v{z}ar, and B. Zupan}
%
\institute{University of Ljubljana, Faculty of Computer and Information Science, Večna pot 113, Ljubljana
\email{\{pavlin.policar,martin.strazar,blaz.zupan\}@fri.uni-lj.si}\\[2pt]
\and
Baylor College of Medicine, Houston, TX 77030, USA
}

@@

Dimensionality reduction techniques, such as t-SNE, can construct informative visualizations of high-dimensional data. When jointly visualising multiple data sets, a straightforward application of these methods often fails; instead of revealing underlying classes, the resulting visualizations expose dataset-specific clusters. To circumvent these batch effects, we propose an embedding procedure that uses a t-SNE visualization constructed on a reference data set as a scaffold for embedding new data points. Each data instance in the secondary data is embedded independently and does not change the reference embedding. This prevents any interactions between instances in the secondary data and implicitly mitigates batch effects. We demonstrate the utility of this approach by analyzing six recently published single-cell gene expression data sets with up to tens of thousands of cells and thousands of genes. The batch effects in our studies are particularly strong as the data comes from different institutions using different experimental protocols. The visualizations constructed by our proposed approach are clear of batch effects, and the cells from secondary data sets correctly co-cluster with cells of the same type from the primary data. We also show the predictive power of our visual classification approach in t-SNE space matches the accuracy of machine learning techniques that consider the entire compendium of features that profile single cells.

\keywords{Batch Effects \and Embedding \and t-SNE \and Visualization \and Single-Cell Transcriptomics \and Data Integration \and Domain Adaptation.}
\end{abstract}

@@

Massive, multi-variate biological data sets often suffer from these
source-specific biases.  The focus of this work is single-cell genomics, a
domain that was selected due to high biomedical relevance and abundance of
recently published data. Single-cell RNA sequencing (scRNA-seq) data sets are
the result of isolating RNA molecules from individual cells, which serve as an
estimate of the expression of cell's genes.  The studies can exceed thousands
of cells and tens of thousands of genes, and typically start with cell type
analysis. Here, it is expected that cells of the same type would cluster
together in two-dimensional data visualization~\cite{seurat}. For instance,
Fig.~\ref{fig:batch_effect}.a shows t-SNE embedded data from mouse brain cells
originating from the visual cortex~\cite{hrvatin2018} and the
hypothalamus~\cite{chen2017}. The figure reveals distinct clusters but also
separates the data from the two brain regions. These two regions share the same
cell types and --- contrary to the depiction in Fig.~\ref{fig:batch_effect}.a
--- we would expect the data points from the two studies to overlap. Batch
effects similarly prohibit the utility of t-SNE in the exploration of
pancreatic cells in Fig.~\ref{fig:batch_effect}.b, which renders the data from
a pancreatic cell atlas~\cite{baron2016} and similarly-typed cells from
diabetic patients~\cite{xin2016}. Just like with data from brain cells,
pancreatic cells cluster primarily by data source, again resulting in a
visualization driven by batch effects.

\begin{figure}[htbp]
  \includegraphics[width=\textwidth]{batch_effect.pdf}
  \caption{Batch effects are a driving factor of variation between the data
  sets. We depict a t-SNE visualization of two pairs of data sets. In each
  pair, the data sets share cell types, so we would expect cells
  from the reference data (blue) to mix with the cells in a secondary data
  sets (orange). Instead, t-SNE clusters data according to the
  data source.}
  \label{fig:batch_effect}
\end{figure}

Current solutions to embedding the data from various data sources address the
batch effect problems up-front. The data is typically preprocessed and
transformed such that the batch effects are explicitly removed. Recently
proposed procedures for batch effect removal include canonical correlation
analysis~\cite{cca} and mutual nearest-neighbors~\cite{mnn,seurat}.  In these
works, batch effects are deemed removed when cells from different sources
exhibit good mixing in a t-SNE visualization.  The elimination of batch effects
may require aggressive data preprocessing which may blur the boundaries between
cell types. Another problem is also the inclusion of any new data, for which
the entire analysis pipeline must be rerun, usually resulting in a different
embedding layout and clusters that have little resemblance to original visualization and
thus require reinterpretation.

We propose a direct solution of rendering t-SNE visualizations to address
batch effects. Our approach treats one of the data sets as a {\em reference}
and embeds the cells from another, {\em secondary data set} to a reference-defined
low-dimensional space.  We construct a t\nobreakdash -SNE embedding using the
reference data set, which is then used as a scaffold to embed the
secondary data. The key idea underpinning our approach is that secondary data
points are embedded independently of one another.

@@

\begin{figure}[htb]
  \includegraphics[width=\textwidth]{transform_brain.pdf}
  \caption{A two-dimensional embedding of a reference containing brain cells
  \textbf{(a)} and the corresponding mapping of secondary data containing hypothalamic
  cells \textbf{(b)}.  The majority of hypothalamic cells were mapped to their
  corresponding reference cluster. For instance, astrocyte cells marked with
  red on the right were mapped to an oval cluster of same-typed cells denoted
  with the same color in the visualization on the left.}
  \label{fig:transform_brain}
\end{figure}

The proposed solution implements a mapping of new data into an existing
t-SNE visualization. While the utility of such an algorithm was already hinted
at in recent publication~\cite{art_of_using_tsne}, we here provide its
practical and theoretically-grounded implementation. Considering the abundance
of recent publications on batch effect removal, we present surprising evidence
that a computationally more direct and principled embedding procedure solves
the batch effects problem when constructing interpretable visualizations from
different data sources.

@@

We describe an end-to-end pipeline that uses fixed t-SNE coordinates as a
scaffold for embedding new (secondary) data, enabling joint visualization of
multiple data sources while mitigating batch effects. Our proposed approach
starts by using t\nobreakdash -SNE to embed a reference data set, with the aim
of constructing a two-dimensional visualization to facilitate interpretation
and cluster classification. Then, the placement of each new sample is optimized
independently via the t\nobreakdash -SNE loss function. Independent treatment
of each data instance from a secondary data set disregards any interactions
present in that data set, and prevents the formation of clusters that would be
specific to the secondary data. Below, we start with a summary of t-SNE and its
extensions (Sec.~\ref{sec:tsne}), introducing the relevant notation, upon which
we base our secondary data embedding approach (Sec.~\ref{sec:transfer}).

\subsection{Data Embedding by t-SNE and Its Extensions\label{sec:tsne}}

@@

When using t-SNE on larger data sets, the standard learning rate $\eta = 200$
has been shown to lead to slower convergence and requires more iterations to
achieve consistent embeddings~\cite{belkina2019automated}. We follow the
recommendation of Belkina \etal ~and use a higher learning rate $\eta = N / 12$
when visualizing larger data sets.

\subsection{Adding New Data Points to Reference Embedding\label{sec:transfer}}

@@

In the optimization step, we refine point positions using batch gradient
descent. We use an adaptive learning rate scheme with momentum to speed up the
convergence, as proposed by Jacobs~\cite{momentum,bh_tsne}. We run gradient
descent with momentum $\alpha$ of $0.8$ for $250$ iterations, where the
optimization converged in all our experiments. The time complexity needed to
evaluate the gradients in Eq.~\ref{eq:gradient} is $\mathcal{O}(N \cdot M)$,
however, by adapting the same polynomial interpolation based approximation,
this is reduced to $\mathcal{O}(\max \{ N, M \})$. The time complexity can further be
reduced to $\mathcal{O}(M)$ by exploiting the fact that the reference embedding remains fixed.

Special care must be taken to reduce the learning rate $\eta$ as the default
value in most implementations ($\eta = 200$) may cause points to ``shoot off''
from the reference embedding. This phenomenon is caused due to the embedding to
a previously defined t-SNE space, where the distances between data points and
corresponding gradients of the optimization function may be quite large. When
running standard t-SNE, points are initialized and scaled to have variance
0.0001. The resulting gradients tend to be very small during the initial phase,
resulting in stable convergence. When embedding new samples, the span of the
embedding is much larger, resulting in substantially larger gradients, and the
default learning rate causes points to move very far from the reference
embedding. In our experiments, we found that decreasing the learning rate to
$\eta \sim 0.1$ produces stable solutions. Alternatively, we can employ gradient
clipping to achieve similar behaviour. This is especially important when
using the interpolation-based approximation, which places a grid of
interpolation points over the embedding space, where the number of grid points
is determined by the span of the embedding. Clearly, if even one point ``shoots
off'' far from the embedding, the number of required grid points may grow
dramatically, increasing the runtime substantially. The reduced learning rate
suppresses this issue, and does not slow the convergence because of the
adaptive learning rate scheme, provided the optimization is run for a
sufficient number of steps.

@@

\vspace{-3mm}
\begin{table}[ht]
\begin{center}
\setlength\tabcolsep{4pt}
\begin{tabular}{l c c r c c}
\toprule
Study & Organism/Tissue & Protocol & Cells & Cell Types & Sparsity (\%) \\
\midrule
Hrvatin \etal & \multirow{2}{*}{mouse brain} & inDrop & 48,266 & 9 & 94 \\
Chen \etal & & Drop-seq & 14,437 & 6 & 93 \\[5pt]
Baron \etal & \multirow{2}{*}{human pancreas} & inDrop & 8,569 & 9 & 91 \\
Xin \etal & & SMARTer & 1,492 & 4 & 86 \\[5pt]
Macosko \etal & \multirow{2}{*}{mouse retina} & Drop-seq & 44,808 & 12 & 97 \\
Shekhar \etal & & Drop-seq & 27,499 & 5 & 96 \\
\bottomrule
\end{tabular}
\end{center}
\caption{Data sets used in our experiments. The first data set in each pair
(Hrvatin \etal, Baron \etal, and Macoscko \etal) was used as a reference.  We
relied on the quality control and annotations from the original publication
and report the number of cell types after preprocessinng.  The cell
annotations were made consistent to annotations from the Cell
Ontology~\cite{cell_ontology}. Notice that different RNA sequencing protocols
were used to estimate gene expressions.}
\label{tab:data sets}
\end{table}

\begin{figure}[htb]
  \includegraphics[width=\textwidth]{transform_pancreas.pdf}
  \caption{Embedding of pancreatic cells from Baron \etal~\cite{baron2016} and
  cells from the same tissue from Xin \etal~\cite{xin2016}. Just like in
  Fig.~\ref{fig:transform_brain}, the vast majority of the cells from the
  secondary data set were correctly mapped to the same-typed cluster of
  reference cells.}
  \label{fig:transform_pancreas}
\end{figure}

\begin{figure}[htb]
  \includegraphics[width=\textwidth]{transform_retina.pdf}
  \caption{An embedding of a large reference of retinal cells from Macosco
  \etal~\cite{macosko2015} \textbf{(a)} and mapping of cells from a smaller study that
  focuses on bipolar cells from Shekhar \etal~\cite{shekhar2016} \textbf{(b)}. We use
  colors consistent with the study by Macosko \etal}
  \label{fig:transform_retina}
\end{figure}

We use a number of additional, recently proposed modifications to enhance the t-SNE visualization of the reference data set.
Kobak and Linderman have shown that the global consistency of embeddings produced by
popular visualization algorithms are largely dependent on their initialization~\cite{kobak2019umap}. By utilizing PCA-based initialization, t-SNE is able to achieve
more meaningful layouts of the resulting clusters (Fig.~\ref{fig:multiscale}.b)
as opposed to using randomly initialized embeddings (Fig.~\ref{fig:multiscale}.a).
Another important extension is the use of multi-scale similarities, which, in addition
to considering short range interactions, also models wider point neighborhoods. Coupled
with PCA-based initializations, this produces even more meaningful visualizations where
clusters form interpretable structures. For instance, consider Fig.~\ref{fig:multiscale}.c, which reveals two meaningful subgroups of GABAergic neurons, corresponding to their
developmental origin, as discussed in Tasic \etal~\cite{tasic2018}, while this division
is less apparent when using PCA-based initialization alone in
Fig.~\ref{fig:multiscale}.b.

In principle, our theoretically-grounded embedding of secondary data into the
scaffold defined by the reference embedding could be simplified with the
application of the nearest neighbors-based procedure. For example, while
describing a set of tricks for t-SNE, Kobak and Berens~\cite{art_of_using_tsne}
proposed positioning new points into a known embedding by placing them in the
median position of their 10 nearest neighbors, where the neighborhood was
estimated in the original data space. Notice that we use this approach as well,
but only for the initialization of positions of new data instances that are
subject to further optimization. Despite both nearest-neighbors search and
t-SNE optimization can be computed in linear time, the former dominates the
runtime (mouse retina example; 44,808 reference, 26,830 secondary cells,
9min NN-search, 13sec optimization).
